Structural Breaks in Inflation Dynamics within the European Monetary Union

### 2011-06-11 pkg/paper/Paper/draft.Rnw
(close to) final touches of Z
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/glogis@117 b679b2f7-6539-45d0-a605-5ddcb976cde9

\author{Thomas Windberger\\Universit\"at Innsbruck \And
        Achim Zeileis\\Universit\"at Innsbruck}
\Plainauthor{Thomas Windberger, Achim Zeileis}

\Abstract{
  In this paper, we test for structural breaks in the inflation dynamics of 21 European countries. To capture
  changes in the first three moments of the inflation rates, we assumed them to be distributed according to 
  a generalized logistic distribution. We find breaks in all but 4 countries.
  Overall, there is a convergence towards a lower inflation rate, but with an increase in variance and a change
  towards negative skewness.
  %% avoid "we"-constructs
  %% "adopt" better than "assume"
}

@@

\Address{
  Thomas Windberger, Achim Zeileis\\
  Department of Statistics\\
  Universit\"at Innsbruck\\
  Universit\"atsstra{\ss}e 15\\
  A-6020 Innsbruck, Austria\\
  E-mail: \email{Thomas.Windberger@student.uibk.ac.at}, \email{Achim.Zeileis@R-project.org} \\
}

@@

\readme{perhaps skip this paragraph?} The way towards the EMU basically consisted of three stages: stage~I
(1990--1994) as a phase of liberalization, stage~II (1995--1998) a phase of
convergence, and stage~III (1999--2001) the transition period, which ended with
the introduction of the euro as legal tender. A possible fourth stage is the
continuing integration of new member states, mostly former communist countries.

Since the beginning of the European Union, the topic of a common currency
was a controversial issue. Although the Economic and Monetary Union (EMU) is now
a fact, the discussion about the economic effects of the euro is far from being
settled. The controversial topics range from the question of whether or not the
eurozone is indeed an optimal currency area \citep[as developed in:][]{Mundell:1961}, 
all the way to the very survival of the euro in light of the budgetary problems of some of its member
states. The effects of monetary unions on a number of macroeconomic indicators,
with inflation being the most important one in this context, is in the center of
an ongoing debate. This concerns the issue of short--run and steady--state
inflation uncertainty -- in context to inflation expectations -- dealt with in
\citet{cap}, or the degree of similarity of short--run dynamic properties of the inflation rates in EU
countries, which is the topic of an investigation by \citet{palomba}.

%Any new state that wants to join the EMU is first obliged to fulfill the
%Maastricht criteria, which, besides setting rules for government debts and
%interest rates, also requires the participation in the ERM~II (Exchange Rate
%Mechanism~II) for two years, while the exchange rates towards the euro are not
%allowed to cross the nominal band.

Throughout the literature, there is still a
considerable degree of uncertainty as to what extent the introduction of the
euro, or monetary unions in general, affect the inflation rate.

However, a deeper inquiry into the effects of monetary unions on the inflation rates of its member
countries is in need of a method able to
capture changes in inflation dynamics. We contribute to this field of research by
developing a new method for testing for structural breaks in the dynamics of the
inflation rates. The framework utilized in this paper allows testing for any 
possible alternations  and thus helps interpreting 
the effects of the EMU. We focus on changes in the first three moments of these 
inflation rates, which we assume to be distributed according to a generalized logistic 
distribution (GL). For this purpose, we build on previous research by \citet{Zeileis+Hornik:2007}.

This \fixme{describe method} framework allows us to assess the evolution of mean, variance, and skewness over time.
Our estimation technique enables the simultaneous estimation of breaks in these moments. 
The skewness parameter turns out to be quite important to measure alterations in the
long run inflation dynamics, as it changes considerably over the time of the
analysis. This trend would be all but invisible using a normal distribution.

\fixme{comment on results here} Overall, we find four periods during which structural breaks
happened. Two of them in the 1990s (early and late), another one in the early 2000s, and the last period covers the financial crisis. There is also a 
tendency towards a lower mean inflation rate, combined with a decreasing difference between
individual inflation rates.

The  remainder of this paper is
structured as follows: Section~\ref{sec:data} presents the data,
Section~\ref{sec:model} presents the model and the estimation techniques used,
Section~\ref{sec:slovenia} illustrates our approach using Slovenia as an
example, Section~\ref{sec:results} presents the results, and
Section~\ref{sec:conclusion} concludes.

\clearpage
\section{Data}\label{sec:data}

@@

\clearpage
\section{Model}\label{sec:model}

@@

The GL distribution in the parameterization employed here is also used in \citet{Shao:2002}
% The type~I GL distribution (also known as skew-logistic distribution), as defined in \citet{Abberger:2002}
%\citet{johnson},
and has the log-density function\footnote{Which is similar to the type~1 GL distribution as defined in \citet{johnson}.}
%
\begin{eqnarray}
%% f(y ~|~ \theta, \sigma, \delta) & = &
%% \frac{\frac{\delta}{\sigma} \cdot \exp\{-\frac{y - \theta}{\sigma}\}}{(1+\exp\{-\frac{y - \theta}{\sigma}\})^{(\delta+1)}} 
\ell(y ~|~ \theta, \sigma, \delta) & = & \log(\delta) - \log(\sigma) - \frac{y - \theta}{\sigma} -
                                         (\delta + 1) \cdot \log\left(1 + \exp\left\{-\frac{y - \theta}{\sigma}\right\}\right)
\label{eq:dglogis}
\end{eqnarray}
%
with location~$\theta$, scale~$\sigma > 0$, and shape~$\delta > 0$. For $\delta = 1$ the
distribution simplifies to the logistic distribution, for $\delta < 1$ or $> 1$ it is
left- or right-skewed, respectively. The corresponding first three moments are
%
\begin{eqnarray}
\mathrm{E}(y) & = & \theta + \sigma (\gamma(\delta) - \gamma(1)) \\
\mathrm{Var}(y)  & = & \sigma^2(\gamma'(\delta)+\gamma'(1)) \\
\mathrm{Skew}(y) & = & \frac{\gamma''(\delta)-\gamma''(1)}{(\gamma'(\delta)+\gamma'(1))^{3/2}}
\end{eqnarray}
%
where $\gamma(\cdot)$, $\gamma'(\cdot)$, and $\gamma''(\cdot)$ are the digamma function and its
first and second derivative, respectively.

@@

\clearpage
\section{Results}\label{sec:results}

@@

To explore the question of convergence, we proceed in two steps:
First, in Section~\ref{sec:bpresults}, the estimation technique proposed
in Section~\ref{sec:model} is applied to each of the countries included in
the data set (see Section~\ref{sec:data}). Second, Section~\ref{sec:aggregate}
aggregates the results from each individual country to reveal different
potential convergence patterns over time within three groups of countries:
core-euro (\Sexpr{paste(colnames(hicps)[euro], collapse = ", ")}),
PIIGS  (\Sexpr{paste(colnames(hicps)[piigs], collapse = ", ")}), and the
rest.

@@

Finally, the inflation rates for the whole euro area \fixme{Is EU euro area or European union?},
as provided by the OECD, are considered and three different regimes are found: (1)~until 
1999 with both a low mean and variance, (2)~the third and final stage of monetary integration,
accompanied with a significant increase in the mean, (3)~a period encompassing
the overheating of the economy that culminated in the financial crisis, showing a big increase
in variance and a lower mean inflation.

\fixme{some concluding remarks about the EU level bridging the current gap to the next subsection}

%There is an increase in mean from
%$\Sexpr{round3(eu_mom[1, 1])}$ (\Sexpr{start(eu)}--\Sexpr{eu_breaks[1]}) to
%$\Sexpr{round3(eu_mom[2, 1])}$ (\Sexpr{eu_breaks[1] + 1/12}--\Sexpr{eu_breaks[2]}) with a slight decrease to
%$\Sexpr{round3(eu_mom[3, 1])}$ (\Sexpr{eu_breaks[2] + 1/12}--\Sexpr{end(eu)}) shortly before the financial crises.
%

@@

\end{tabular}
\caption{\label{tab:results2} Aggregation of estimated moments from Table~\ref{tab:results1} across countries in
January~\Sexpr{yrs[1]}, \Sexpr{yrs[2]}, \Sexpr{yrs[3]}, and \Sexpr{yrs[4]}. Dots signal that the moments have not
changed from the previous time point, i.e., that no breakpoint was estimated in the corresponding time period.}
\end{center}
\end{table}

\fixme{intro needed}

\fixme{also ``zoom out'' more, a big picture needs to be established}

Table~\ref{tab:results2} presents lower quartile,
median, and upper quartile (25\%, 50\%, 75\% percentiles) of the first three moments of the
inflation series in January \Sexpr{yrs[1]}, \Sexpr{yrs[2]}, \Sexpr{yrs[3]} and \Sexpr{yrs[4]}. The countries are grouped
into the ``core'' of the eurozone, the PIIGS countries  and all the others. For this, we took account of possible
breaks in the moments.

The time periods are chosen according to four fundamental change periods. Namely, the early 1990s, with breaks in
Austria, Finland, France, Sweden, and UK. The late 1990s, with breaks in Greece, Italy, Luxembourg, Portugal, Czech Republic,
Estonia, and Hungary. And also the period after the second phase of convergence, with breaks in the early 2000s (Poland, Slovakia, and Slovenia)
and the two countries with the most visible effect of the financial crisis (Ireland and Spain).

In the core-zone, there is a decrease in the lower quartile from \Sexpr{yrs[1]} to
\Sexpr{yrs[2]}. However, the interquartile range decreased by almost 20\%. The variance increased
slightly, but also with a 15\% decrease in interquartile range. The skewness
stayed symmetric in the median, with almost an 8\% increase in the interquartile range.

The PIIGS countries significantly decreased the median inflation rates from \Sexpr{yrs[1]} to \Sexpr{yrs[4]}. 
Whereas the interquartile range was very tight in \Sexpr{yrs[1]}, it increased in \Sexpr{yrs[2]} and again in \Sexpr{yrs[4]}.
However, the median values are much smaller. The variance increased from \Sexpr{yrs[1]} to \Sexpr{yrs[2]} but did not change much 
in \Sexpr{yrs[4]}. The interquartile range did not change much too. Median skewness changed considerable 
from highly positive to negative values. This is mostly due to the changes associated with the financial crisis, that was
responsible for deflationary shocks. The interquartile range of skewness decreased considerably.

The other countries were successful in decreasing all three moments from \Sexpr{yrs[1]} to \Sexpr{yrs[3]}. Still, their median
inflation rate is higher than those of the core countries. Also, the interquartile range of the mean decreased considerably. 
Contrary to the euro--zone and the PIIGS countries, variance decreased also, both in terms of the median value and the
interquartile range. The skewness decreased as well, as did the interquartile range of the skewness parameter.

In context of the convergence issue, we find that, with the interquartile range of the mean of all the aggregated zones declining and getting tighter, there is  
evidence of a convergence towards some common mean inflation rate in the countries of our sample, both within the subgroups and the three groups put together.
This is also visible in the median values, that  diverge much less than at the beginning in \Sexpr{yrs[1]}.

@@

\clearpage
\section{Conclusion}\label{sec:conclusion}

\readme{start with 1) short recap of what we did, 2) conclusions}

To assess changes in the dynamics of the inflation rates of a number of European countries both within
and without the EMU, we developed a new method for estimating structural breaks resting on the 
assumption that the inflation rates are GL distributed.

The results suggest that a GL distribution is appropriate for the modeling of the seasonally 
adjusted inflation rate series and that failing to account for a change in the skewness of the 
distribution might lead to wrong conclusions concerning the dynamics.

The results of \citet{holte}, whose simulations indicated that the standard deviation of the home 
CPI inflation rate can be substantially reduced by joining a monetary union, can not be empirically validated.

For the countries in the core of the eurozone, one clearly visible trend is the rise in 
inflation volatility in almost half of the euro countries. \citet{emerson} emphasize  the fact that a high inflation rate is also more
variable and uncertain and thus causes more relative price variability, leading
to a less efficient price mechanism. From an economic point of view, a very volatile inflation rate will 
likely  contribute to greater macroeconomic instability. In the literature, there is no consensus about the effects of higher 
inflation volatility on the inflation rate. \citet{grier} find no evidence 
that higher inflation uncertainty raises the average inflation rate -- at least in the case of the USA. However,
\cite{jaro} finds a positive correlation between the level and the standard deviation of inflation. Naturally, an
efficient price mechanism will benefit from inflation rates that do not diverge too much from the expectations.

The good news is that mean inflation declined considerably, most notably in the Eastern European countries that clearly 
benefited from the monetary discipline that is demanded for those countries who want to join the EMU.

\clearpage
\section*{Computational details}

@@

@@

>>>>

\author{Thomas Windberger\\Universit\"at Innsbruck \And
        Achim Zeileis\\Universit\"at Innsbruck}
%% \author{\hfill Thomas Windberger \hfill Achim Zeileis\thanks{Correspondence
%% to: Achim Zeileis, Department of Statistics,  Universit\"at Innsbruck,
%% Universit\"atsstr.~15, A-6020 Innsbruck, Austria. Tel: +43/512/507-7103. Fax: +43/512/507-2851.
%% E-mail: \email{Achim.Zeileis@R-project.org}} \hfill \hfill \\
%% {\small \it Department of Statistics, Universit\"at Innsbruck, Austria}}
\Plainauthor{Thomas Windberger, Achim Zeileis}

\renewcommand{\abstractname}{Summary}
\Abstract{
  To assess the effects of the EMU on inflation rate dynamics of
  its member states, the inflation rate series for 21~European countries
  are investigated for structural changes. To capture changes in mean,
  variance, and skewness of inflation rates, a generalized logistic
  model is adopted and complemented with structural break tests and
  breakpoint estimation techniques. These reveal considerable differences
  in the patterns of inflation dynamics and the structural changes
  therein. Overall, there is a convergence towards a lower mean inflation rate
  with reduced skewness, but accompanied by an increase in variance.
}

@@

\Address{
  Thomas Windberger, Achim Zeileis\\
  Department of Statistics\\
  Universit\"at Innsbruck\\
  Universit\"atsstra{\ss}e 15\\
  A-6020 Innsbruck, Austria\\
  E-mail: \email{Thomas.Windberger@student.uibk.ac.at}, \email{Achim.Zeileis@R-project.org} \\
  Telephone: +43/12/507-7103 \\
  Fax: +43/512/507-2851 \\
}

@@

Since the beginning of the European Union (EU), the topic of a common currency
was a controversial issue. Although the Economic and Monetary Union (EMU) is now
a fact, the discussion about the economic effects of the euro is far from being
settled. The controversial topics range from the question of whether or not the
eurozone is indeed an optimal currency area \citep[as developed in][]{Mundell:1961},
all the way to the very survival of the euro in light of
the budgetary problems of some of its member states. The effects of monetary
unions on a number of macroeconomic indicators, with inflation being of increased
importance, is at the center of an ongoing debate. This
concerns the issue of short-run and steady-state inflation uncertainty -- in relation
to inflation expectations -- as dealt with in \citet{cap}, or the degree of
similarity of short-run dynamic properties of the inflation rates in EU
countries, which is the topic of the investigation by \citet{palomba}. 
Throughout the literature, there is still a considerable degree of uncertainty
as to what extent the introduction of the euro, or monetary unions in general,
affects the inflation rate.

As the way towards the EMU essentially consisted of three stages (I:~1990--1994,
a phase of liberalization; II: 1995--1998, a phase of convergence; III: 1999--2001,
a transition period ending with the introduction of the euro as legal tender),
it seems natural to ask whether this transition was accompanied by structural changes
in the inflation dynamics in the different European countries. Furthermore, a
possible fourth stage is the continuing integration of new member states, mostly
former communist countries. Typically, changes in the mean level of the
inflation rates are of most interest but there are also reasons for investigating
changes in higher moments, such as variance and skewness. \citet{emerson} emphasize that
a high inflation rate is usually also more variable and uncertain and thus
causes more relative price variability, leading to a less stability and a less
efficient price mechanism. Furthermore, \cite{jaro} finds a positive correlation
between the level and the standard deviation of inflation.

%% From an economic point of view, a very volatile inflation rate will 
%% likely  contribute to greater macroeconomic instability. In the literature,
%% there is no consensus about the effects of higher  inflation volatility on the
%% inflation rate. \citet{grier} find no evidence that higher inflation
%% uncertainty raises the average inflation rate -- at least in the case of the
%% USA. Naturally, an efficient price mechanism will
%% benefit from inflation rates that do not diverge too much from the expectations.

%% Any new state that wants to join the EMU is first obliged to fulfill the
%% Maastricht criteria, which, besides setting rules for government debts and
%% interest rates, also requires the participation in the ERM~II (Exchange Rate
%% Mechanism~II) for two years, while the exchange rates towards the euro are not
%% allowed to cross the nominal band.

Therefore, a thorough investigation of the evolution of the inflation dynamics
within a country over time requires assessment not only of the mean level but
also higher moments. Hence, we contribute to this field of research by
developing a new method for testing for structural breaks based on a generalized
logistic (GL) distribution. This provides flexible modeling of means, variances,
and skewness, while also allowing for fat tails. Structural breaks are tested
using score-based tests \citep{Andrews:1993,Zeileis:2005} and breakpoints are estimated
by extending the least squares approach of \cite{bp03} to (quasi-)maximum likelihood
estimation of GL distributions \citep[similar in spirit to the Gaussian model
employed by][]{Zeileis+Shah+Patnaik:2010}.

The empirical results show that there are different patterns and timings of
structural breaks in the inflation dynamics of different groups of countries.
Inflation rates in the core euro area did not change significantly after
completion of the first convergence stage, yielding low mean and variance.
The Southern euro area member states (and Ireland), on the other hand, managed
to decrease the mean level of inflation in the mid-1990s, but at the price
of higher volatility. Finally, the Eastern European countries reduced both mean
and variance on their way towards joining the EMU. Generally, the inflation rates
converged from very different distributions at the beginning of the sample to
rather similar patterns at the end of the 2000s. Skewness also plays an
important role in this convergence: Some Southern and Eastern countries
have clear breaks from right-skewed to roughly symmetric inflation rates,
signalling a reduction of size and frequency of outlying high inlation rates.

The  remainder of this paper is
structured as follows: Section~\ref{sec:data} briefly presents the data,
Section~\ref{sec:model} introduces the model and the estimation techniques used,
%% Section~\ref{sec:slovenia} illustrates our approach using Slovenia as an example, 
Section~\ref{sec:results} discusses the empirical results, and
Section~\ref{sec:conclusion} concludes.

\section{Data}\label{sec:data}

@@

\section{Model}\label{sec:model}

@@

The GL distribution employed here corresponds to the type~1 GL distribution 
from\citet{johnson}, using the parametrization and inference as in \citet{Shao:2002}. It 
has the log-density function
%
\begin{eqnarray}
%% f(y ~|~ \theta, \sigma, \delta) & = &
%% \frac{\frac{\delta}{\sigma} \cdot \exp\{-\frac{y - \theta}{\sigma}\}}{(1+\exp\{-\frac{y - \theta}{\sigma}\})^{(\delta+1)}} 
\ell(y ~|~ \theta, \sigma, \delta) & = & \log(\delta) - \log(\sigma) - \frac{y - \theta}{\sigma} -
                                         (\delta + 1) \cdot \log\left(1 + \exp\left\{-\frac{y - \theta}{\sigma}\right\}\right)
\label{eq:dglogis}
\end{eqnarray}
%
with location~$\theta$, scale~$\sigma > 0$, and shape~$\delta > 0$. For $\delta = 1$ the
distribution simplifies to the logistic distribution, for $\delta < 1$ or $> 1$ it is
left- or right-skewed, respectively. The corresponding first three moments are
%
\begin{eqnarray}
\mathrm{E}(y) & = & \theta + \sigma (\gamma(\delta) - \gamma(1)) \\
\mathrm{Var}(y)  & = & \sigma^2(\gamma'(\delta)+\gamma'(1)) \\
\mathrm{Skew}(y) & = & \frac{\gamma''(\delta)-\gamma''(1)}{(\gamma'(\delta)+\gamma'(1))^{3/2}}
\end{eqnarray}
%
where $\gamma(\cdot)$, $\gamma'(\cdot)$, and $\gamma''(\cdot)$ are the digamma function and its
first and second derivative, respectively.

@@

\section{Results}\label{sec:results}

@@

To explore the question of convergence, we proceed in two steps:
First, in Section~\ref{sec:bpresults}, the estimation technique proposed
in Section~\ref{sec:model} is applied to each of the countries included in
the data set (see Section~\ref{sec:data}). Second, Section~\ref{sec:aggregate}
aggregates the results from each individual country to reveal different
potential convergence patterns over time within three groups of countries:
core-euro (\Sexpr{paste(colnames(hicps)[euro], collapse = ", ")}),
PIIGS  (\Sexpr{paste(colnames(hicps)[piigs[c(4, 2, 3, 1, 5)]], collapse = ", ")}), and the
other countries (\Sexpr{paste(gsub("CzechRepublic", "Czech Republic", colnames(hicps))[others], collapse = ", ")}).

@@

Finally, the inflation rates for the whole euro area\footnote{This is the
euro area 17~countries series as provided by the \citet{oecd},
starting in \Sexpr{start(eu) - 1/12}. It is calculated as a weighted average of the HICPs
of now seventeen euro countries.} are considered and
three different regimes are found: (1)~until 1999 with both a low mean and
variance, (2)~the third and final stage of monetary integration, accompanied
by a significant increase in the mean, (3)~a period encompassing the
overheating of the economy and its culmination in the financial crisis, showing a
large increase in variance and a lower mean inflation.

@@

\end{tabular}
\caption{\label{tab:results2} Aggregation of estimated moments from Table~\ref{tab:results1} across countries in
Jan~\Sexpr{yrs[1]}, \Sexpr{yrs[2]}, \Sexpr{yrs[3]}, and \Sexpr{yrs[4]}. Dots signal that the moments have not
changed from the previous time point, i.e., that no breakpoint was estimated in the corresponding time period.}
\end{center}
\end{table}

Given the results above for the individual countries, a natural question is
whether there is convergence on an aggregate level and, if so, which pattern it
follows. Hence, we aggregate the results from Table~\ref{tab:results1} for the
individual countries into three groups and at four time points to reflect the
general dynamics of European inflation rates during the time period of interest.
The groups have been selected based on prior expectations (i.e., not based on
the breakpoints estimation results) as the ``core'' of the euro zone
(\Sexpr{paste(colnames(hicps)[euro], collapse = ", ")}), the PIIGS countries,
and the rest. Similarly, the time points have been chosen as to fall ``between''
times where transitions in one or more groups of countries could be expected:
Jan~\Sexpr{yrs[1]} during stage~I on the way towards the EMU, Jan~\Sexpr{yrs[2]}
during stage~II, Jan~\Sexpr{yrs[3]} after transition into the EMU but before the
financial crisis, and Jan~\Sexpr{yrs[4]} after the crisis. The result of these
considerations is presented in Table~\ref{tab:results2} which extracts the three
moments for each country from Table~\ref{tab:results1} at the indicated time
points and aggregates them within the group. Lower quartile, median, and upper
quartile (25\%, 50\%, 75\% percentiles) are provided to assess both changes in
location and spread across countries.

For the core euro countries, there are only changes from stage~I to stage~II
because after \Sexpr{yrs[2]} no breaks have been detected. The changes in this
first phase did not affect mean inflation much, while variance increased
slightly (in both quartiles) and skewness somewhat decreased (from slightly
right- to slightly left-skewed). The heterogeneity (assessed by the
interquartile range) decreased for all three moments of the inflation rates.
This reflects that the monetary policies of the corresponding countries were
already closely connected prior to stage~I and now even moved somewhat closer
together.

For the PIIGS countries, changes occured from \Sexpr{yrs[1]} to
\Sexpr{yrs[2]} and then again during the crisis, before \Sexpr{yrs[4]}.
In the first phase, mean inflation was almost halved, accompanied by a
doubling of variance and a change from clearly right-skewed to approximately
symmetric. The latter reflects that the occurence of shocks with high inflation
rates have been successfully reduced. Moreover, the changes in the PIIGS
countries between \Sexpr{yrs[3]} and \Sexpr{yrs[4]} lead to a further
decrease in mean inflation and associated skewness which are both caused by
several extremely low inflation (or even deflation) rates during the time
following the crisis.

The other countries do not exhibit any significant breakpoints during the
mid-1990s, but many changes occur in the transition period from \Sexpr{yrs[2]}
to \Sexpr{yrs[3]}, encompassing both the completion of transformation from
planned to free market economies and the intended entry into the EMU.
During this time, the countries were successful in
decreasing all three moments of their inflation distributions while
simultaneously decreasing the heterogeneity (as captured by the corresponding
interquartile ranges) across countries. Before the transition all three
moments were extremely large compared to the other two groups, while afterwards
all moments are much closer to those of the other groups. In Jan~\Sexpr{yrs[3]},
mean inflation is in between the levels for the core euro and PIIGS countres,
variance is still slightly higher, and skewness is not too far from symmetry.

In context of the convergence issue, we find that up to \Sexpr{yrs[3]}
the inflation dynamics of all three groups of countries become much
more similar, both within and across groups. The interquartile ranges
of the means decline and become tighter for all groups and there appears to
be convergence towards some common mean inflation rate in the
countries of our sample. Variances also move more closely together, corresponding
to an increase for the core euro and PIIGS countries but a clear decrease
for the others. Right-skewness (and thus large outlying inflation rates)
has been reduced for all three groups of countries but to different extents.
Thus, while ending in similar inflation dynamics, the timing and patterns of
changes in this convergence process are rather different across
the three groups, depending on the different starting situations of the
countries.

@@

\section{Conclusion}\label{sec:conclusion}

To assess changes in the dynamics of the inflation rates of a number of European countries
-- both within and outside the EMU -- a new method for estimating structural breaks in
inflation rates is proposed. The method allows to test for changes in mean, variance,
and skewness of inflation rate series by adopting a (quasi-)likelihood model based on the generalized
logistic (GL) distribution. Thus, unlike normally distributed models, this approach
can deal with fat tails and (potentially changing) skewness.

The empirical results suggest that a GL distribution is appropriate for modeling seasonally 
adjusted inflation rate series and is able to uncover important changes in all three moments.
Towards the end of the 2000s, before the financial crisis, our results broadly confirm
convergence hypotheses as the first three moments from all countries in our sample
move closer together.
Especially the changes in skewness -- caused by a reduction of size and frequency of outlying high inflation
rates in some countries -- might have gone unnoticed with traditional structural change techniques,
leading to potentially misleading conclusions concerning the dynamics.

At the beginning of the sample, many PIIGS and Eastern countries did have very high inflation
rates, while for the core euro countries, convergence was already achieved in the mid-1990s.
The PIIGS countries lowered their inflation rates considerably in preparation to the adoption
of the euro. The Eastern European countries lowered their high initial inflation level in the
late 1990s or early 2000s, prior to the entry into the ERM~II. They clearly benefited from the
monetary discipline that is demanded for those countries who want to join the EMU.

\section*{Computational details}

@@

\section*{Acknowledgments}

We are grateful to Jesus Crespo Cuaresma for comments and discussions about
an earlier version of the manuscript, and also to the participants of the workshop
of the Research Platform ``Empirical and Experimental Economics'' for stimulating discussions,
leading to substantial improvements of the paper.

@@

\end{document}
\SweaveOpts{eval=FALSE}
